---
title: "COMETS Analytics Single Cohort Analysis in R"
date: "`r Sys.Date()`"
output: html_document
---

# This sample Rmd file provides a walkthrough of the basics of using COMETS Analytics in R
# You can visit the COMETS Analytics vignette at: https://github.com/CBIIT/R-cometsAnalytics/blob/master/RPackageSource/README.md
# Sample input sheet file publicly available at: https://github.com/CBIIT/R-cometsAnalytics/blob/master/InputRepository/cometsInputAge.xlsx

```{r setup, include=FALSE}

if (!require("devtools")) {
    install.packages("devtools")
}

library(devtools)
devtools::install_github("CBIIT/R-cometsAnalytics/RPackageSource@master")

library(RcometsAnalytics)

```

# 1. readCOMETSinput()

## Read input sheet and perform integrity checks

```{r load example data}

# Retrieve the full path of the input data
dir <- system.file("extdata", package="RcometsAnalytics", mustWork=TRUE)
csvfile <- file.path(dir, "cometsInputAge.xlsx")

# Read in and process the input data
exmetabdata <- RcometsAnalytics::readCOMETSinput(csvfile)

```

## Plot the variance distribution of the metabolite values

```{r plot var}

RcometsAnalytics::plotVar(exmetabdata,titlesize=12)

```

## Plot the distribution of the number of samples with minimum values for each metabolite

```{r plot min}

# note: minimum values are defined as those with the minimum across all samples (ignoring missing values)

RcometsAnalytics::plotMinvalues(exmetabdata,titlesize=12)

```


# 2. getModelData()

## Set up pre-specified models based on input sheet

```{r pre-spec setup}

# set up "1 Age" model as specified in the input sheet
exmodeldata <- RcometsAnalytics::getModelData(exmetabdata,modlabel="1 Age") 

```

## Set up interactive mode to specify in R

```{r interactive setup}

# set up metabolites ~ age model using interactive mode
exmodeldata2 <- RcometsAnalytics::getModelData(exmetabdata, modelspec="Interactive",
    exposures=c("age"))

```


# 3. runModel()

## Run pre-specified models based on input sheet

```{r run pre-spec model}

excorrdata <- RcometsAnalytics::runModel(exmodeldata, exmetabdata, "test_cohort")
RcometsAnalytics::OutputListToExcel(filename="test_cohort_corr1.xlsx", excorrdata)
RcometsAnalytics::showModel(excorrdata,nlines=3)

```

## Run interactive model

```{r run interactive model}

excorrdata2 <- RcometsAnalytics::runModel(exmodeldata2, exmetabdata, "test_cohort")
RcometsAnalytics::OutputListToExcel(filename="test_cohort_corr2.xlsx", excorrdata2)
RcometsAnalytics::showModel(excorrdata2,nlines=3)

```


# 4. Example of another analysis

## Set up and run stratified correlation analaysis

```{r stratified model example}

exmodeldata2 <- RcometsAnalytics::getModelData(exmetabdata,modelspec="Interactive",
                   outcomes=c("lactose","lactate"),
                exposures=c("age","bmi_grp"),strvars="race_grp")

excorrdata2  <- RcometsAnalytics::runModel(exmodeldata2,exmetabdata,"test_cohort")

RcometsAnalytics::showModel(excorrdata2,nlines=3)

```


# 5. Example of super-batch mode

## Use super-batch mode to run all models as specified in input sheet

```{r run all models}

exallmodels <- RcometsAnalytics::runAllModels(exmetabdata, cohortLabel = "test_cohort") # whatever cohortLabel you choose will be used in autosaved files of results

RcometsAnalytics::showModel(exallmodels,nlines=3)

```

