---
title: "Running COMETS Analytics locally"
author: "Ewy Mathé, Ella Temprosa, Bill Wheeler"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
vignette: >
  %\VignetteIndexEntry{Running COMETS Analytics locally}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

COMETS Analytics support all cohort-specific analyses of the COMETS consortium. This collaborative work is done via the COMETS harmonization group activities. For more information, see the [COMETS website] (http://epi.grants.cancer.gov/comets/).

## Data Input Format

The required input file shoud be in excel format with the following 6 sheets: 

  1. Metabolites - from harmonized metabolites output
  2. SubjectMetabolites - abundances in columns and subject in rows
  3. SubjectData - other exposure and adjustment variables
  4. VarMap - maps the variables needed to conduct the cohort specific analysis. Specify the name of variables under CohortVariable column. if the VarReference has the same name in the cohort, leave blank
  5. Models - models used to conduct COMETS analysis. Outcomes, exposures and adjustment can specify multiple covariates delimited by spaces (ie: age bmi).
  6. ModelOptions - optional sheet containing model specific options

An example input file is available [**HERE**](https://www.comets-analytics.org/examples/cometsInput.xlsx).

## Analysis Workflow for correlation analysis

### 1. Load Data

The first step of the analysis is to load in the data with the *readCOMETSinput()* function.  Input for this function is an Excel spreadsheet, per the description above.  

```{r}
# Retrieve the full path of the input data
dir <- system.file("extdata", package="COMETS", mustWork=TRUE)
csvfile <- file.path(dir, "cometsInputAge.xlsx")

# Read in and process the input data
exmetabdata <- COMETS::readCOMETSinput(csvfile)
```

To plot some the distribution of variances for each metabolite:
```{r, eval=T}
COMETS::plotVar(exmetabdata,titlesize=12)
```

To plot the distribution of minimum/missing values:
```{r, eval=T}
COMETS::plotMinvalues(exmetabdata,titlesize=12)
```

### 2. Get Model Data

There are 2 ways to specify your model, batch or interactive.
In Batch mode, models are specified in your input file Models sheet. 
The model information needs to be read in with the function *getModelData()* and 
processed so the software knows which models to run.
The following call defines the "1 Gender adjusted" model from the Models
sheet in the input file to be run.

```{r}
exmodeldata <- COMETS::getModelData(exmetabdata,modlabel="1 Gender adjusted")
```
In Interactive mode, models are specified as parameters. 
The model information needs to be read in with the function *getModelData()* and 
processed so the software knows which models to run.  
The following call defines the model with age and bmi_grp as the exposure variables,
 and includes only the subjects with age > 40 and bmi_grp > 2.
```{r}
exmodeldata2 <- COMETS::getModelData(exmetabdata, modelspec="Interactive",
	exposures=c("age","bmi_grp"), where=c("age>40","bmi_grp>2"))
```

### 3. Run Simple Unstratified Correlation Analysis
The *runModel()* function is the main function for running a single model, and by
default, a correlation analysis is performed. 
```{r}
excorrdata  <- COMETS::runModel(exmodeldata2,exmetabdata,"DPP")
```

The output of the correlation analysis can then be compiled and output to an Excel file
 with the following function:
```{r, eval=F}
COMETS::OutputListToExcel(filename="DPP_corr.xlsx", excorrdata)
```

To view the first 3 lines of the correlation analysis output, simply type:
```{r}
COMETS::showModel(excorrdata,nlines=3)
```

To display the heatmap of the resulting correlation matrix, use the showheatmap function.
```{r in-text-fig, eval=T}
COMETS::showHeatmap(excorrdata)
```
\
\
\
To display the hierarchical clustering of the resulting correlation matrix, 
use the showHClust function. 
This diplay requires at least 2 rows and 2 columns in the correlation matrix.

```{r, eval=T}
exmodeldata<-COMETS::getModelData(exmetabdata,modelspec = "Interactive",exposures = c("bmi_grp","age"))
excorrdata  <- COMETS::runModel(exmodeldata,exmetabdata,"DPP")
COMETS::showHClust(excorrdata, showticklabels=FALSE)
```

Results can be written to an output Excel file with the following command:

```{r eval=F}
COMETS::OutputListToExcel("Model1.xlsx", excorrdata)
```

### 4. Run Stratified Correlation Analysis
A stratified correlation analysis can be performed by specifiying stratification
variables in the call to *getModelData()*. If more than one stratification variable
is specified, then the strata will be defined by all unique combinations of the
stratification variables. The following call will define a model stratified 
by race_grp.
```{r}
  exmodeldata2 <- COMETS::getModelData(exmetabdata,modelspec="Interactive",
                   outcomes=c("lactose","lactate"),
	            exposures=c("age","bmi_grp"),strvars="race_grp")
```

The stratified correlation analysis is run by calling the *runModel()* function.          
```{r}
  excorrdata2  <- COMETS::runModel(exmodeldata2,exmetabdata,"DPP")
```

### 5. Linear regression with lm
Call *getModelData()* to define a model which adjusts for age group, 
has lactose and lactate as outcome variables, and has age and bmi group 
as the exposure variables.
```{r}
  exmodeldata <- COMETS::getModelData(exmetabdata,modelspec="Interactive", adjvars="age_grp",
                   outcomes=c("lactose","lactate"), exposures=c("age","bmi_grp"))
```
To run a linear regression using the lm function, a list of options must be passed into
*runModel()* with the model option set to "lm".
```{r}
  lm_results  <- COMETS::runModel(exmodeldata, exmetabdata, "DPP", op=list(model="lm"))
  print(lm_results)
```

### 6. Linear regression with glm
Run a linear regression using the glm function for the same variables as above.
The default family used with glm is "gaussian", which corresponds to a
linear regression.
The Effects data frame will be the same as with lm, but the ModelSummary data frame
will contain some different columns.
```{r}
  glm_results  <- COMETS::runModel(exmodeldata, exmetabdata, "DPP", op=list(model="glm"))
  print(all.equal(lm_results$Effects, glm_results$Effects))
  print(glm_results$ModelSummary)
```

### 7. Logistic regression with glm
Call *getModelData()* to define a model which adjusts for age group, 
has nested_case as the outcome variable, and has lactose and lactate
as the exposure variables. The variable nested_case must be a binary 0-1 variable.
```{r}
  exmodeldata <- COMETS::getModelData(exmetabdata,modelspec="Interactive", adjvars="age_grp",
                   outcomes="nested_case", exposures=c("lactose","lactate"))
```
To run a logistic regression, the list of options op must also include a model.options 
list with family set to "binomial".
```{r}
  op <- list(model="glm", model.options=list(family="binomial"))
  glm_results  <- COMETS::runModel(exmodeldata, exmetabdata, "DPP", op=op)
  print(glm_results)
```

### 8. Poisson regression with glm
Call *getModelData()* to define a model which adjusts for age group, 
has n_visits as the outcome variable, and has lactose and lactate
as the exposure variables. The variable n_visits must be a non-negative integer valued variable.
```{r}
  exmodeldata <- COMETS::getModelData(exmetabdata,modelspec="Interactive", adjvars="age_grp",
                   outcomes="n_visits", exposures=c("lactose","lactate"))
```
To run a Poisson regression, the list of options op must also include a model.options 
list with family set to "poisson".
```{r}
  op <- list(model="glm", model.options=list(family="poisson"))
  poisson_results  <- COMETS::runModel(exmodeldata, exmetabdata, "DPP", op=op)
  print(poisson_results)
```

### 9. Run Analysis on all models defined in the input Excell sheet ('super-batch' mode)

All models desginated in the input file can be run with one command, 
and individual output Excel files or correlation results will be written
 in the current directory by default. The function returns a list of objects.

```{r eval=F}
 allresults <- COMETS::runAllModels(exmetabdata,writeTofile=F)

```{r}
sessionInfo()
```
