---
title: "Running COMETS Analytics locally"
author: "Ewy Mathé, Ella Temprosa"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

COMETS Analytics support all cohort-specific analyses of the COMETS consortium. This collaborative work is done via the COMETS harmonization group activities. For more information, see the [COMETS website] (http://epi.grants.cancer.gov/comets/).

## Data Input Format

The required input file shoud be in excel format with the following 5 sheets: 

  1. Metabolites - from harmonized metabolites output
  2. SubjectMetabolites - abundances in columns and subject in rows
  3. SubjectData - other exposure and adjustment variables
  4. VarMap - maps the variables needed to conduct the cohort specific analysis. Specify the name of variables under CohortVariable column. if the VarReference has the same name in the cohort, leave blank
  5. Models - models used to conduct COMETS analysis. Outcomes, exposures and adjustment can specify multiple covariates delimited by spaces (ie: age bmi).
  6. Options - optional sheet containing model specific options

An example input file is available [**HERE**](https://www.comets-analytics.org/examples/cometsInput.xlsx).

## Analysis Workflow for correlation analysis

### 1. Load Data

The first step of the analysis is to load in the data with the *readCOMETSinput()* function.  Input for this function is an Excel spreadsheet, per the description above.  

```{r}
# Retrieve the full path of the input data
dir <- system.file("extdata", package="COMETS", mustWork=TRUE)
csvfile <- file.path(dir, "cometsInputAge.xlsx")

# Read in and process the input data
exmetabdata <- COMETS::readCOMETSinput(csvfile)
```

To plot some the distribution of variances for each metabolite:
```{r, eval=T}
COMETS::plotVar(exmetabdata,titlesize=12)
```

To plot the distribution of minimum/missing values:
```{r, eval=T}
COMETS::plotMinvalues(exmetabdata,titlesize=12)
```

### 2. Get Model Data

There are 2 ways to specify your model, batch or interactive.
In Batch mode, models are specified in your input file. The model information needs to be read in with the function *getModelData()* and processed so the software knows which models to run.  Input for this function is the data input in the previous step:

```{r}
exmodeldata <- COMETS::getModelData(exmetabdata,modlabel="1 Gender adjusted")
```
In Interactive mode, models are specified as parameters. The model information needs to be read in with the function *getModelData()* and processed so the software knows which models to run.  Input for this function is the data input in the previous step:

```{r}
exmodeldata <- COMETS::getModelData(exmetabdata, modelspec="Interactive",
	colvars=c("age","bmi_grp"), where=c("age>40","bmi_grp>2"))
```

### 3. Run Simple Correlation Analysis

The unstratified correlation analysis is run by calling the function *runModel()*.  This function runs the model(s) that is(are) defined in the input data (Models tab).  

```{r}
excorrdata  <- COMETS::runModel(exmodeldata,exmetabdata,"DPP")
```

The output of the correlation analysis can then be compiled and output to an Excel file with the following function:
```{r, eval=F}
COMETS::OutputListToExcel(filename="DPP_corr.xlsx", excorrdata)
```

To view the first 3 lines of the correlation analysis output, simply type:
```{r}
COMETS::showCorr(excorrdata,nlines=3)
```

To display the heatmap of the resulting correlation matrix, use the showheatmap function.
```{r in-text-fig, eval=T}
COMETS::showHeatmap(excorrdata)
```
\
\
\
To display the hierarchical clustering of the resulting correlation matrix, use the showHClust function. This diplay requires at least 2 rows and 2 columns in the correlation matrix.

```{r, eval=T}
exmodeldata<-COMETS::getModelData(exmetabdata,modelspec = "Interactive",colvars = c("bmi_grp","age"))
excorrdata  <- COMETS::runModel(exmodeldata,exmetabdata,"DPP")
COMETS::showHClust(excorrdata)
```

Results can be written to an output Excel file with the following command:

```{r eval=F}
COMETS::OutputListToExcel("Model1.xlsx", excorrdata)
```

### 4. Run Stratified Correlation Analysis

The stratified correlation analysis is run by calling the function *runModel()*.  This function runs the model(s) that is(are) defined in the input data (Models tab) or in interactive mode. In this example, exmodeldata includes an object scovs that specifies the stratification variable.        

```{r}
  exmodeldata2 <- COMETS::getModelData(exmetabdata,modelspec="Interactive",rowvars=c("lactose","lactate"),
	colvars=c("age","bmi_grp"),strvars="race_grp")
  excorrdata2  <- COMETS::runModel(exmodeldata2,exmetabdata,"DPP")
```

### 5. Linear regression
Run a linear regression using the lm function adjusting for age group with age and bmi group as the exposures.
```{r}
  exmodeldata <- COMETS::getModelData(exmetabdata,modelspec="Interactive", adjvars="age_grp",
                   rowvars=c("lactose","lactate"), colvars=c("age","bmi_grp"))
  lm_results  <- COMETS::runModel(exmodeldata, exmetabdata, "DPP", op=list(model="lm"))
  print(lm_results)
```

### 6. Linear regression with glm
Run a linear regression using the glm function for the same variables as above.
The Effects data frame will be the same as with lm, but the ModelSummary data frame
will contain some different columns.
```{r}
  glm_results  <- COMETS::runModel(exmodeldata, exmetabdata, "DPP", op=list(model="glm"))
  print(all.equal(lm_results$Effects, glm_results$Effects))
```
### 7. Logistic regression with glm
Run a logistic regression using the glm function with the binary variable nested_case
as the outcome. Here we must specify model.options in the op list to ensure the
correct family is used.
```{r}
  exmodeldata <- COMETS::getModelData(exmetabdata,modelspec="Interactive", adjvars="age_grp",
                   rowvars="nested_case", colvars=c("lactose","lactate"))
  op <- list(model="glm", model.options=list(family="binomial"))
  glm_results  <- COMETS::runModel(exmodeldata, exmetabdata, "DPP", op=op)
```

### 8. Run Analysis on all models defined in the input Excell sheet ('super-batch' mode)

All models desginated in the input file can be run with one command, and individual output Excel files or correlation results will be written in the current directory by default. The function returns a list of objects.

```{r eval=F}
 allresults <- COMETS::runAllModels(exmetabdata,writeTofile=F)

```{r}
sessionInfo()
```
