---
title: "Running meta-analyses with RcometsAnalytics locally"
author: "Ewy Mathé, Ella Temprosa, Bill Wheeler"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
vignette: |
  %\VignetteIndexEntry{Running RcometsAnalytics locally} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

## Introduction

RcometsAnalytics supports all cohort-specific analyses and meta-analyses of the R COMETS Analytics consortium. 
This collaborative work is done via the COMETS Data Infrastructure interest group activities. 
For more information, see the COMETS [**website**](http://epi.grants.cancer.gov/comets/).
This vignette demonstrates using the RcometsAnalytics R package to run a meta-analysis 
from the command line assuming that the user does not already have model output files.
The first three steps are similar to steps found in the vignette for running models locally:
[**Comets_vignette**](https://cbiit.github.io/R-cometsAnalytics/cometsvignette_v2.1.html).

## Meta-analysis Background Information
A meta-analysis is performed by combining the results from more than one cohort in order to 
increase the power of association tests. Each set of cohort results should be from the
same underlying model, and each cohort should be independent from all other
cohorts, that is, no overlapping subjects between any two cohorts. 
This package will conduct both fixed-effects and random-effects meta-analyses based
on the DerSimonian-Laird method.

## RcometsAnalytics R package
The main functions for performing meta-analyses in the RcometsAnalytics R package are 
*runMeta()* and *runAllMeta()*. Both functions require that the model output from 
each cohort has been saved to a file. The function *meta_calc()* is the core function 
that performs the calculations.
Documentation of the RcometsAnalytics R package can be found here
  [**manual**](https://github.com/CBIIT/R-cometsAnalytics/blob/gh-pages/RcometsAnalytics-manual-v3.0.pdf).

## Data Input Format
The files input to the *runMeta()* meta-analysis function contain the output from the *runModel()* function.
These files can be Excel files with extension .xlsx or R object files with extension .rda.
Each file should contain the results from a single model run from one of the cohorts.

## Example Workflow with two cohorts
Let "cohort_1" and "cohort_2" be the names of the two cohorts. For each cohort, we will run the 
same model using the *runModel()* function, save the model output, and then run a meta-analysis
using the *runMeta()* function.

### 1. Load Data 
For each cohort, load the cohort's data with the *readCOMETSinput()* function.

```{r}
# Retrieve the full paths of each input data
dir   <- system.file("extdata", package="RcometsAnalytics", mustWork=TRUE)
file1 <- file.path(dir, "cohort_1.xlsx")
file2 <- file.path(dir, "cohort_2.xlsx")

# Read in and process the input data
data1 <- RcometsAnalytics::readCOMETSinput(file1)
data2 <- RcometsAnalytics::readCOMETSinput(file2)
```

### 2. Get Model Data

Get the model data for each cohort. The model must be the same for all cohorts.
The model to be run will have age as a continuous exposure variable, all metabolites
as the outcome variables, and bmi as a categorical adjustment variable.

```{r}
modeldata1 <- RcometsAnalytics::getModelData(data1, modelspec="Interactive", 
                          adjvars="bmi_grp", outcomes=NULL, exposures="age")
modeldata2 <- RcometsAnalytics::getModelData(data2, modelspec="Interactive", 
                          adjvars="bmi_grp", outcomes=NULL, exposures="age")
```

### 3. Linear Regression
The *runModel()* function will first be called to obtain the results from each
cohort for a linear regression analysis. Since the meta-analysis function *runMeta()*
requires files for input, the results from *runModel()* will be saved to files.
The output file names defined below have the specific format 
<model name>__<cohort name>__<date>.ext, where ext is "xlsx" or "rda". 
This format of the file name is important when using the *runAllMeta()*
function shown in section 6. 
The model name is specified as "AgeAdjustedForBMI".
```{r}
outfile1  <- paste0(getwd(), "/", "AgeAdjustedForBMI__cohort_1__", Sys.Date(), ".xlsx")
outfile2  <- paste0(getwd(), "/", "AgeAdjustedForBMI__cohort_2__", Sys.Date(), ".xlsx")
results1  <- RcometsAnalytics::runModel(modeldata1, data1, "cohort_1", 
             op=list(model="lm"))
results2  <- RcometsAnalytics::runModel(modeldata2, data2, "cohort_2", 
             op=list(model="lm"))
tmp       <- RcometsAnalytics::OutputListToExcel(outfile1, results1)
tmp       <- RcometsAnalytics::OutputListToExcel(outfile2, results2)
```

### 4. Meta Analysis of Results
Run the meta-analysis with the *runMeta()* function, and then display
the first two rows of the results. In this example, a vector of file names
is passed into the *runMeta()* function; however a list of file information
can be passed in as we will see in a later example below. See the *runMeta()*
function in the user 
[**manual**](https://github.com/CBIIT/R-cometsAnalytics/blob/gh-pages/RcometsAnalytics-manual-v3.0.pdf)
for complete documentation.
```{r}
ret <- RcometsAnalytics::runMeta(c(outfile1, outfile2))
ret$Results[1:2,]
```

### 5. Example using runAllMeta
The *runAllMeta()* function is for running meta-analyses in super-batch mode.
The function takes three arguments, with the first argument being a character
vector of file names and/or folder names. The second argument is the output
folder to write the results to, and the third argument is an optional file 
for meta-analysis options. The result files "outfile1" and "outfile2" created
in section 3 above will be used for this meta-analysis. Instead of passing
in the file names, the folder name where these files reside (the working directory)
 will be passed in. The output from *runAllMeta()* will also be written
to the working directory.
```{r}
tmp <- RcometsAnalytics::runAllMeta(getwd(), getwd())
```

The output files names created by *runAllMeta()* will be of the form
\<model name\>\_\_meta\_\_\<date\>.ext. 
```{r}
res.file <- paste0(getwd(), "/AgeAdjustedForBMI__meta__", Sys.Date(), ".xlsx")
```

Read in the table of results and compare to the previous meta-analysis performed 
in section 4. The results should be the same.
```{r}
res <- as.data.frame(readxl::read_excel(res.file, "Metaresults"))
all.equal(ret$Results, res)
```

Delete the files created
```{r}
file.remove(c(outfile1, outfile2, res.file))
```

```{r}
#sessionInfo()
```
