---
title: "Running meta-analyses with RcometsAnalytics locally"
author: "Ewy Mathé, Ella Temprosa, Bill Wheeler"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
vignette: |
  %\VignetteIndexEntry{Running RcometsAnalytics locally} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

## Introduction

RcometsAnalytics supports all cohort-specific analyses and meta-analyses of the R COMETS Analytics consortium. 
This collaborative work is done via the COMETS Data Infrastructure interest group activities. 
For more information, see the COMETS [**website**](http://epi.grants.cancer.gov/comets/).
This vignette demonstrates using the RcometsAnalytics R package to run a meta-analysis 
from the command line assuming that the user does not already have model output files.
The first three steps are similar to steps found in the vignette for running models locally:
[**Comets_vignette**](https://cbiit.github.io/R-cometsAnalytics/cometsvignette_v2.1.html).

## Meta-analysis Background Information
A meta-analysis is performed by combining the results from more than one cohort in order to 
increase the power of association tests. Each set of cohort results should be from the
same underlying model, and each cohort should be independent from all other
cohorts, that is, no overlapping subjects between any two cohorts. 
This package will conduct both fixed-effects and random-effects meta-analyses based
on the DerSimonian-Laird method.

## RcometsAnalytics R package
The main functions for performing meta-analyses in the RcometsAnalytics R package are 
*runMeta()* and *runAllMeta()*. Both functions require that the model output from 
each cohort has been saved to a file. The function *meta_calc()* is the core function 
that performs the calculations.
Documentation of the RcometsAnalytics R package can be found here
  [**manual**](https://github.com/CBIIT/R-cometsAnalytics/blob/gh-pages/RcometsAnalytics-manual.pdf).

## Data Input Format
The files input to the *runMeta()* meta-analysis function contain the output from the *runModel()* function.
These files can be Excel files with extension .xlsx or R object files with extension .rda.
Each file should contain the results from a single model run from one of the cohorts.

## Example Workflow with two cohorts
Let "cohort_1" and "cohort_2" be the names of the two cohorts. For each cohort, we will run the 
same model using the *runModel()* function, save the model output, and then run a meta-analysis
using the *runMeta()* function.

### 1. Load Data 
For each cohort, load the cohort's data with the *readCOMETSinput()* function.

```{r}
# Retrieve the full paths of each input data
dir   <- system.file("extdata", package="RcometsAnalytics", mustWork=TRUE)
file1 <- file.path(dir, "cohort_1.xlsx")
file2 <- file.path(dir, "cohort_2.xlsx")

# Read in and process the input data
data1 <- RcometsAnalytics::readCOMETSinput(file1)
data2 <- RcometsAnalytics::readCOMETSinput(file2)
```

### 2. Get Model Data

Get the model data for each cohort. The model must be the same for all cohorts.
The model to be run will have age as a continuous exposure variable, all metabolites
as the outcome variables, and bmi as a categorical adjustment variable.

```{r}
modeldata1 <- RcometsAnalytics::getModelData(data1, modelspec="Interactive", 
                          adjvars="bmi_grp", outcomes=NULL, exposures="age")
modeldata2 <- RcometsAnalytics::getModelData(data2, modelspec="Interactive", 
                          adjvars="bmi_grp", outcomes=NULL, exposures="age")
```

### 3. Correlation Analysis
The *runModel()* function will first be called to obtain the results from each
cohort for a correlation analysis. Since the meta-analysis function *runMeta()*
requires files for input, the results from *runModel()* will be saved to files.
The files *runMeta()* accepts have the extension ".xlsx" or ".rda".  
```{r}
outfile1  <- paste0(getwd(), "/", "corr_results1.xlsx")
outfile2  <- paste0(getwd(), "/", "corr_results2.rda")
results1  <- RcometsAnalytics::runModel(modeldata1, data1, "cohort_1", out.file=outfile1)
results2  <- RcometsAnalytics::runModel(modeldata2, data2, "cohort_2", out.file=outfile2)
```

### 4. Meta Analysis of Correlation Results
Run the meta-analysis with the *runMeta()* function, and then display
the first five rows of the results.
```{r}
ret <- RcometsAnalytics::runMeta(c(outfile1, outfile2))
ret$Results[1:5,]

# Delete the results files from each cohort
file.remove(c(outfile1, outfile2))
```


```{r}
sessionInfo()
```
